Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-27 article.tex
spellings

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is a high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough for being able to detect
interactions, and that it would therefore be preferrable to lower their
importance by changing appropriatly their magnitude. Additionnally, it is
well-known that neurons may also spike because of a global high activity
\citep{stetter2012model}. In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
\end{align}
which effect is to magnify the importance of spikes that occur in case of low
global activity (measured by $\sum_{j} x^{t}_j$), as observed for instance
around $t=4\text{s}$ in the figure. Note that the particular case where there
is no activity, i.e., $\sum_{j} x^{t}_j = 0$, is solved by setting $w(x^{t}_i)
= 1$.

@@

Our procedure to infer connections between neurons first assumes that
the (filtered) fluorescence concentrations of all $p$ neurons at each
time point can be modeled as a set of random variables $X = \{X_1,
\dots, X_p\}$ that are idependently drawn from the same time invariant
joint probability distribution $P_X$. %% Or, say otherwise, we assume
%% that there exists a probability distribution $P_X$ such that the
%% fluorescence concentrations at a particular time $t$ can be seen as an
%% independent realization of $P_X$.
As a consequence, our inference method does not exploit the time
ordering of the observations (although time ordering is exploited by
the filters).

@@

Finally, let us remark that performance of our simple method shows to
be quite sensitive to the values of parameters (e.g., $f_1$ or $f_2$
or the value of the threshold $\tau$) in the combined function of the
filtering and infering processes. One approach, further referred
to as \textit{Averaged Partial correlation} statistics, for improving
its robustness is to average correlation statistics over various
values of the parameters, thereby reducing the variance of its
predictions. Further details about parameter selection are provided in
Appendix~\ref{app:optimized}.

@@

In the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalized transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:curves}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer of the
DREAM5 challenge \citep{marbach2012}. When transposed to neural networks, this
method uses as a confidence score for the edge going from neuron $i$ to neuron
$j$ the importance score of variable $X_i$ in a Random Forest model trying to
predict $X_j$ from all variables in $X\setminus X_j$. To reduce the
computational cost of this method, we had however to limit each tree in the
Random Forest model to a maximum depth of 3. This constraint potentially
severly affects the performance of this method with respect to the use of fully
grown trees. We nevertheless provide these results for comparison purpose. PC
and GENIE3 were both applied on the same pre-filtered time series as partial
correlation ($h\circ g \circ f_1$). For GENIE3, we built 10000 trees per neuron
and we used default settings for all other parameters (except for the maximal
tree depth). For GTE, we reproduced the exact same setting (conditioning level
and pre-processing) that was used for the challenge.

>>>>

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is a high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough for being able to detect
interactions, and that it would therefore be preferable to lower their
importance by changing appropriately their magnitude. Additionally, it is
well-known that neurons may also spike because of a global high activity
\citep{stetter2012model}. In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
\end{align}
which effect is to magnify the importance of spikes that occur in case of low
global activity (measured by $\sum_{j} x^{t}_j$), as observed for instance
around $t=4\text{s}$ in the figure. Note that the particular case where there
is no activity, i.e., $\sum_{j} x^{t}_j = 0$, is solved by setting $w(x^{t}_i)
= 1$.

@@

Our procedure to infer connections between neurons first assumes that
the (filtered) fluorescence concentrations of all $p$ neurons at each
time point can be modeled as a set of random variables $X = \{X_1,
\dots, X_p\}$ that are independently drawn from the same time invariant
joint probability distribution $P_X$. %% Or, say otherwise, we assume
%% that there exists a probability distribution $P_X$ such that the
%% fluorescence concentrations at a particular time $t$ can be seen as an
%% independent realization of $P_X$.
As a consequence, our inference method does not exploit the time
ordering of the observations (although time ordering is exploited by
the filters).

@@

Finally, let us remark that performance of our simple method shows to
be quite sensitive to the values of parameters (e.g., $f_1$ or $f_2$
or the value of the threshold $\tau$) in the combined function of the
filtering and inferring processes. One approach, further referred
to as \textit{Averaged Partial correlation} statistics, for improving
its robustness is to average correlation statistics over various
values of the parameters, thereby reducing the variance of its
predictions. Further details about parameter selection are provided in
Appendix~\ref{app:optimized}.

@@

In the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalized transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:curves}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer of the
DREAM5 challenge \citep{marbach2012}. When transposed to neural networks, this
method uses as a confidence score for the edge going from neuron $i$ to neuron
$j$ the importance score of variable $X_i$ in a Random Forest model trying to
predict $X_j$ from all variables in $X\setminus X_j$. To reduce the
computational cost of this method, we had however to limit each tree in the
Random Forest model to a maximum depth of 3. This constraint potentially
severely affects the performance of this method with respect to the use of fully
grown trees. We nevertheless provide these results for comparison purpose. PC
and GENIE3 were both applied on the same pre-filtered time series as partial
correlation ($h\circ g \circ f_1$). For GENIE3, we built 10000 trees per neuron
and we used default settings for all other parameters (except for the maximal
tree depth). For GTE, we reproduced the exact same setting (conditioning level
and pre-processing) that was used for the challenge.
